Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-29 paper.tex
Updates in abstract and introduction

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate 6
  architectures with metrics measuring integration challenge,
  robustness, extensibility, scalability and other specific
  features. Results show the pros and cons of the different
  architectures: tight integration and task encapsulation are the
  simplest architectures and are therefore the easiest to integrate
  and the most robust ones, most architectures perform the same in
  terms of extensibility, the pool model is the most scalable and
  meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for software architects on
  how to integrate science gateways and workflow engines.
\end{abstract}

@@

Workflow engines are critical components in the ecosystem of tools and
services offered by science gateways for the efficient and transparent
execution of applications on distributed infrastructures. Several
software architectures can be adopted to integrate workflow engines in
science gateways, with important consequences on the required
development effort and on the resulting system. This paper describes,
exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework that highlight the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, we evaluate the architectures with
metrics measuring integration challenge, robustness, extensibility,
scalability and other specific features. The paper closes on a
discussion and conclusion that compare architectures and highlight
recommendations for science gateway architects.

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. For instance, FSL can launch tasks on SGE through
its fsl\_sub tool, Nipype has execution plugins for S/OGE, Torque,
IPython, ssh and multi-processor servers, and PSOM is interfaced to
Torque, SGE, MOAB, LSF, and HTCondor. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways may be tailored to a particular
engine that manages critical functions such as task execution, data
transfers and authentication. Other gateways may be more general and
host applications executed by different types of engines. Finally,
some science gateways may not use any workflow engine and focus on the
execution of simple tasks -- this class of systems is not covered by
our analysis.

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.

Scalability is also a crucial concern, since science gateways are
multi-user systems that manage substantial amounts of data and execute
computationally-intensive analyses. To this end, science gateways
may include different instances of the same engine type and implement
some load-balancing policy among them. They may also use advanced
task scheduling policies on the infrastructure, e.g. to ensure fairness
among users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a highly desirable characteristic in production gateways. In this respect, the
multiplicity of software components and interactions involved in architectures is
a challenge as it often leads to complex systems.
Architectures showing reduced system complexity will greatly
facilitate the implementation effort towards robust interactions
which, in turn, have a positive impact on characteristics such as
gateway predictability, transparence, general reliability and results
traceability and reproducibility; characteristics that are also key to
a good user experience. \closednote{Marc-Etienne}{I find that
  something is weak in this last paragraph. Robustness will help with
  transparency, sure, but it seems only a small facet.  Also, the text
  seem to timidly favor architectures with less
  components/interactions without much explanation... Complex systems
  can be robust, but at greater costs...  Here is one: "Architectures
  showing reduced system complexity will greatly facilitate the
  implementation effort towards robust interactions which, in turn,
  have a positive impact on characteristics such as gateway
  predictability, transparence, general reliability and results
  traceability and reproducibility; characteristics that are also key
  to a good user experience."}{Tristan}{Agreed, fixed as suggested.}

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble science gateways. The main
frameworks are Apache Airavata~\cite{marru2011apache}, the Catania
Science Gateway Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateway instances
were built using such
frameworks~\cite{kacsuk2014science,ardizzone2012decide}. Other gateway
instances were also built independently from such specialized
framework, using regular web development tools and libraries to access
distributed systems. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13} are
two such gateways.

\subsection{Infrastructures}

In this paper, infrastructure consists of the computing and storage
resources involved in the workflow execution, as well as the software
services used to access these resources. Infrastructures can be composed of
computing or file servers, databases, clusters, grids or clouds. Some
workflow engines and science gateways may assume specific
characteristics about the infrastructure, such as the presence of a
shared file system between the computing nodes, the availability of a
global task meta-scheduler, the presence of a file catalog, etc. In
the analysis presented below, such specific requirements are not
discussed. Instead, we consider the infrastructure as an abstract
system that can execute tasks and store data regardless of the
enabling mechanisms.

>>>>

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We describe and evaluate 6
  architectures with metrics measuring integration challenge,
  robustness, extensibility, scalability and other specific
  features. Results show the pros and cons of the different
  architectures: tight integration and task encapsulation are the
  simplest architectures and are therefore the easiest to integrate
  and the most robust ones, most architectures perform the same in
  terms of extensibility, the pool model is the most scalable and
  meta-workflows are only available in nested workflows and workflow
  import. Our analysis provides insights for science gateway
  architects and workflow engine developers.
\end{abstract}

@@

Workflow engines are critical components in the ecosystem of tools and
services offered by science gateways for the efficient and transparent
exploitation of distributed infrastructures. Several
software architectures can be adopted to integrate workflow engines in
science gateways, with important consequences on the 
development effort required and on the resulting system. This paper describes,
exemplifies and compares such architectures, based on
system-independent representations of their main components and
interactions.

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework that highlight the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, we evaluate the architectures with
metrics measuring integration challenge, robustness, extensibility,
scalability and other specific features. The paper closes on a
discussion and conclusion that compare architectures and highlight
recommendations for science gateway architects and workflow engine developers.

@@

At the same time, toolboxes were emerging in various scientific
domains to facilitate the assembly of software components in
consistent ``pipelines''. In neuroimaging, our primary domain of
interest, tools such as Nipype (Neuroimaging in Python, Pipelines and
Interfaces~\cite{gorgolewski2011nipype}), PSOM (Pipeline System for
Octave and Matlab~\cite{bellec2012pipeline}), SPM (Statistical
Parametric Mapping~\cite{ashburner2011spm}) and FSL (FMRIB Software
Library~\cite{Jenkinson2012782}) provide abstractions and functions to
handle the data and computing flow between processes implemented in a
variety of programming languages. Such tools were interfaced to
computing infrastructures, in particular clusters\closednote{Tristan}{Reverted from Marc's 'cluster scheduler'}{}{}, to execute tasks at
a high throughput. Some of these tools also support
advanced features such as provenance tracking (Nipype, PSOM),
redundancy detection across analyses to avoid re-computation (PSOM),
and so on. A wide-array of workflows have been implemented using these
pipelines and are now shared across neuroimaging groups
world-wide. These domain-specific engines represent a tremendous
opportunity for science gateways to leverage existing tools and
applications. They nicely complement e-Science engines that are more
oriented towards the exploitation of distributed computing
infrastructures, in particular grids and clouds.

@@

Integration between workflow engines and science gateways varies
across systems. Some science gateways are tailored to a particular
engine while other ones are more general and host applications
executed by different types of engines. Some gateways may not even use any
workflow engine and focus on the execution of simple tasks.

Extensibility is an important property of the integration, as new
workflows are added frequently, different types and versions of
workflow engines may be integrated over time, and different kinds of
infrastructure can be targeted.Scalability is also a crucial concern
for such multi-user systems that manage substantial amounts of data
and execute computationally-intensive analyses. To this end, science
gateways may balance the load among different instances of the same
engine. They may also use advanced task scheduling policies on the
infrastructure, e.g. to ensure fairness among
users~\cite{FERR-14,CPE:CPE3708} or to improve
fault-tolerance~\cite{wilkins2008teragrid,FERR-13}.

Robustness is a highly desirable characteristic as well. Architectures
showing reduced system complexity will greatly facilitate the
implementation effort towards robust interactions which, in turn, have
a positive impact on characteristics such as gateway predictability,
transparence, general reliability and results traceability and
reproducibility; characteristics that are also key to a good user
experience. \closednote{Marc-Etienne}{I find that something is weak in
  this last paragraph. Robustness will help with transparency, sure,
  but it seems only a small facet.  Also, the text seem to timidly
  favor architectures with less components/interactions without much
  explanation... Complex systems can be robust, but at greater
  costs...  Here is one: "Architectures showing reduced system
  complexity will greatly facilitate the implementation effort towards
  robust interactions which, in turn, have a positive impact on
  characteristics such as gateway predictability, transparence,
  general reliability and results traceability and reproducibility;
  characteristics that are also key to a good user
  experience."}{Tristan}{Agreed, fixed as
  suggested.}\note{Tristan}{Work on the this sentence again.}

@@

Various science gateways have been developed, including frameworks
used as building blocks to assemble other systems, for instance Apache
Airavata~\cite{marru2011apache}, the Catania Science Gateway
Framework~\cite{Ardizzone2012} and
WS-PGRADE/gUSE~\cite{Kacsuk2012}. Numerous science gateways were built
using such frameworks~\cite{kacsuk2014science,ardizzone2012decide},
and other ones, e.g. CBRAIN~\cite{SHER-14} and VIP~\cite{GLAT-13},
were built without any framework.

\subsection{Infrastructure}

In this paper, infrastructure consists of the computing and storage
resources involved in workflow execution, as well as the software
services used to access these resources. Infrastructure can be
composed of computing or file servers, databases, clusters, grids or
clouds. Some workflow engines and science gateways may assume specific
characteristics, such as the presence of a shared file system between
the computing nodes, the availability of a global task meta-scheduler,
the presence of a file catalog, etc. In the analysis presented below,
such specific requirements are not discussed. Instead, we consider the
infrastructure as an abstract system that can execute tasks and store
data regardless of the enabling mechanisms.
